Amicable Contempt

### 2023-12-22 manuscript/manuscript.qmd
Substantial edits to theory section

@@

@@

I use an adaptation of negotiation theory [@Raiffa:1982; @Mayer:2010] to conceptualize this dynamic relationship between governments and INGOs. I represent the interaction between regimes and INGOs in any given issue area (human rights, development, advocacy, etc.) in simple two-dimensional policy space (see @fig-theory-full). Each actor has a preference for a certain level of advocacy or service provision, marked as solid circles in the figure. Governments set *de jure* policies to prevent INGO activities from going beyond the regime's ideal point, while INGOs will engage in programming that, when possible, fits within the bounds set by the regime. Each actor has a range of flexibility in the implementation of their regulations or their programming, shown in the horizontal lines flanking the ideal points. The space where the two ranges overlap determines both the regime's *de facto* level of restrictions and the INGO's actual mix of programming in that country and constitutes the practical operating environment for the INGO.

@@

A regime's preferences comprise two components: (1) formal *de jure* legislation, represented by the ideal point, and (2) *de facto* implementation of those policies, represented by the negotiation space around the ideal point. Regimes establish a regulatory environment for international civil society that ranges in permissiveness—some countries allow for a large degree of contentious INGO activity, while others are highly restrictive and provide little room for dissonance. In turn, and analogous to the regime's preferences, an INGO's negotiation preferences comprise two components: (1) the organization's normative principles and ideals, represented by the ideal point, (2) the organization's programmatic flexibility in actually implementing those principles, represented by the negotiation space. INGOs have a desire to engage in a certain level of advocacy, generally defined by their mission, vision, and values. INGOs with more instrumental flexibility—that is, a diverse menu of programs, professionalized staff, consistent revenue streams, and collaborative relationships with governments, foundations, and other NGOs—can fine tune their programs to fit the regulatory environments they face, while those with less flexibility are less able to make such adjustments.

The interaction between the actors' negotiation positions introduces two additional dynamics that are crucial for determining how authoritarian regimes regulate INGOs and how well INGOs can respond correspondingly. First, the distance between the two parties ideal points determines the contentiousness of their relationship. INGOs that engage in programming close to the regime's preferences are naturally more cooperative, either providing services that the government needs or advocating for issues that are unlikely to lead to political instability. Accordingly, INGOs that are more cooperative will be *compliant* with the legal limits imposed by the regime. Organizations that pursue more contentious programming further from the regime's ideal point, on the other hand, will have an *antagonistic* relationship and will be more likely to openly resist the restrictions imposed on them.

@@

The relationship between INGOs and regimes is dynamic, and shifts in INGO flexibility and de facto regime civil society regulations can create different practical operating environments, demonstrated in @tbl-theory-grid. For instance, there are two circumstances where INGOs will feel less restricted. First, as seen in the top left quadrant of @tbl-theory-grid, if the distance between the two parties' ideal points is short and the negotiation spaces overlap—as with INGOs providing non-contentious programming like disaster relief, humanitarian services, or education—INGOs will be cooperative and face a tolerant operating environment. Because of preference alignment, these INGOs will not need to rely on their flexibility to maintain access to their host countries.

More contentious organizations with preferences that diverge from the regime's ideal point, on the other hand, can also face a tolerant environment as long as they have a large amount of programmatic flexibility (see bottom left quadrant of @tbl-theory-grid). For example, @Heiss:2019a explores how human rights INGOs advocate for democratization, political reform, and increased respect for human rights by relying on networks of field offices, funded by dozens of private foundations and government grants—resources that give these organizations the flexibility to remain in their host countries and maintain a fairly tolerant operating environment.

Likewise, there are two circumstances where INGOs will feel restricted, or even expelled. Most often, organizations that face restricted operating environments engage in contentious programming and work in countries where regimes have limited de facto regulatory environments, leaving little room for negotiation or overlap (see bottom right quadrant of @tbl-theory-grid). In contrast, INGOs that are naturally less contentious and more prone to compliance can face expulsion if they lack flexibility or it the civil society regulatory environment is too constrained (see top right quadrant of @tbl-theory-grid). For instance, TODO in 2015, less contentious INGOs like Mercy Corps were restricted after India's de facto environment constricted [@Carothers:2015]. No formal laws were changed, but the amount of leeway granted to INGOs shrunk considerably and Modi's government froze assets and cancelled registrations of hundreds of INGOs.

>>>>

::: {.content-visible when-format="html"}
# Introduction {.hide}
:::

@@

@ChristensenWeinstein:2013; @DupuyRonPrakash:2015; @DupuyRonPrakash:2016; @BakkeMitchellSmidt:2020

@@

@@

I conceptualize this dynamic relationship between governments and INGOs using an adaptation of negotiation theory [@Raiffa:1982; @Mayer:2010]. I represent the interaction between regimes and INGOs in any given issue area (human rights, development, advocacy, etc.) in simple two-dimensional policy space (see @fig-theory-full). Each actor has a preference for a certain level of contentiousness in INGO advocacy or service provision. Regime preferences comprise two components: (1) formal *de jure* legislation, marked as a point, and (2) *de facto* implementation of those policies, represented by the space around the ideal point. Regimes establish a regulatory environment for international civil society that ranges in permissiveness—some countries allow for a large degree of contentious INGO activity, while others are highly restrictive and provide little room for dissonance.

In turn, INGO preferences comprise two components: (1) the organization's normative principles and ideals, represented the point, and (2) the organization's programmatic flexibility in actually implementing those principles, represented by the space around the point. INGOs have a desire to engage in a certain level of contentiousness programming, defined by their mission, vision, and values. INGOs with more programmatic flexibility—that is, a diverse menu of programs, professionalized staff, consistent revenue streams, and collaborative relationships with governments, foundations, and other NGOs—can fine tune their programs to fit the regulatory environments they face, while those with less flexibility are less able to make such adjustments.

The interaction between the actors' positions introduces two additional dynamics that are crucial for determining how governments regulate INGOs and how well INGOs can respond. First, the distance between the two parties' ideal points determines the nature of their relationship. INGOs that engage in programming close to the regime's preferences are naturally more cooperative, either providing services that the government needs or advocating for issues that are unlikely to lead to political instability. Accordingly, INGOs that are more cooperative will be *cooperative* with the legal limits imposed by the regime. Organizations that pursue more contentious programming further from the regime's ideal point, on the other hand, will have an *antagonistic* relationship and will be more likely to openly resist the restrictions imposed on them.

@@

The relationship between INGOs and regimes is dynamic, and shifts in INGO flexibility and de facto regime civil society regulations can create different practical operating environments, demonstrated in @tbl-theory-grid. For instance, INGOs can enjoy a tolerant practical operating environment when facing both cooperative and antagonistic relationships with the host government. The environment will be tolerant and cooperative when the distance between the two parties' ideal points is small and the negotiation spaces overlap, as with INGOs like the Red Cross and Oxfam, which provide non-contentious programming like disaster relief or humanitarian services. More antagonistic organizations with preferences that diverge from the regime's ideal point, on the other hand, can also face a tolerant environment as long as (1) the regime provides substantial openness and freedom for advocacy when enforcing its laws, and (2) the organizations have a large amount of flexibility in pursuing their programs. For instance, advocacy organizations like Amnesty International or Reporters Without Borders often target democratic governments that violate human rights but do not face expulsion from those countries because of robust associational laws that permit such advocacy. Because of the alignment of the practical operating environment, INGOs in both situations are tolerated by their host governments and maintain access to the country.

At the core of the phenomenon of closing civic space is the fact that INGOs can feel restricted—or even expelled—regardless of whether their relationship with the host government is cooperative or antagonistic. INGOs that have a naturally antagonistic relationship with their host governments—for instance, human rights organizations targeting authoritarian states—can face a limited (or even nonexistent) practical operating environment if the de facto civil society regulatory environment is highly constricted. These organizations can still work under restrictive conditions as long as they can rely on their programmatic flexibility. For example, @Heiss:2019a explores how human rights INGOs in the Middle East advocate for democratization, political reform, and increased respect for human rights by relying on networks of field offices, funded by dozens of private foundations and government grants—resources that give these organizations the flexibility to remain in their host countries and maintain a narrow practical operating environment. INGOs that (1) lack this flexibility, or that (2) hope to work in countries with overly constricted civil society regulations—face expulsion. In contrast, INGOs that have a naturally more cooperative relationship with their host governments can also face restrictions if they lack flexibility or it the civil society regulatory environment becomes too constrained. For instance, in the wake of shifts in domestic political dynamics, non-contentious humanitarian INGOs like Mercy Corps were restricted after India's de facto environment constricted in 2015 [@Carothers:2015]. No formal laws were changed, but the amount of leeway granted to INGOs shrunk considerably as Narendra Modi's government froze assets and cancelled registrations of hundreds of INGOs.
